\chapter{\label{chap:techniques}Details on the available techniques}
In this Chapter, more details on special options and the theory behind each technique is given.
The techniques marked with a star (*) are implemented with OpenMP and MPI options (see \cite{Sardjan_2020}), while techniques implemented with OpenMP options are marked with a plus (+).
All other techniques are implemented for a single CPU.
Running these techniques on a cluster reserving multiple nodes or CPUs will be waste of time.
When running them interactively a warning may be produced, which can in that case be ignored.

\section{Analyse}
The Hamiltonian is analysed and different statistical properties are provided including the average delocalization size of Thouless \cite{Thouless.1974.PR.13.93}.
This inverse participation ratio (IPR) is expressed as:
\begin{equation}
	D_{IPR}=\left\langle\frac{1}{N}\sum_i\left(\sum_j |c_{ij}|^{4}\right)^{-1}\right\rangle.
\end{equation}
The Manhattan exciton size is calculated as \cite{Jansen.2025.J.Chem.Phys..162.074113}:
\begin{equation}
	D_{MES}=\left\langle\frac{1}{N}\sum_i\left(\sum_j |c_{ij}|\right)^{2}\right\rangle.
\end{equation}
In a file named \ilc{Analyse.dat} statistics is given for each site including.
This include the average energy and the standard deviation.
The average coupling strength (signed sum of all couplings of a given site with all other sites) and the standard deviation of this quantity.
In a file named \ilc{Av\_Hamiltonian.dat} the average Hamiltonian is stored as a single snapshot in the GROASC format.
Furthermore, the density matrix is calculated for the spectral region defined by the upper and lower bounds of \ilk{MinFrequencies} and \ilk{MaxFrequencies}.
This is done both with and without a weight determined by the contribution to the absorption spectrum as:
\begin{equation}
\rho_{ij}=\sum_k \Big\langle c_{ik}^* c_{jk}  \Theta(\omega_{k}-\omega_{min})\Theta(\omega_{max}-\omega_k)\Big\rangle
\end{equation}
Where $c_{jk}$ is the wavefunction coefficient of eigenstate $k$ on site $j$ and $\omega_k$ is the wavenumber associated with that eigenstate.
The brackets symbolize the ensemble average over the trajectory.
\begin{equation}
\rho^{\textrm{spectral}}_{ij}=\sum_k \Big\langle |\vec{\mu}_k|^2 c_{ik}^* c_{jk}  \Theta(\omega_{k}-\omega_{min})\Theta(\omega_{max}-\omega_k)\Big\rangle
\end{equation}
Here $\vec{\mu}_k$ is the transition-dipole vector of eigenstate $k$.
This weighted density matrix, thus, emphasizes the states contributing to the absorption spectrum in the given spectral window.
The absolute value density matrix defined as:
\begin{equation}
	\rho^{\textrm{ADM}}_{ij}=\sum_k \Big\langle |c_{ik}^* c_{jk}|  \Theta(\omega_{k}-\omega_{min})\Theta(\omega_{max}-\omega_k)\Big\rangle,
\end{equation}
reveal information about the delocalization in the system.
The participation ration matrix defined as:
\begin{equation}
	\rho^{\textrm{PRM}}_{ij}=\sum_k \Big\langle (c_{ik}^* c_{jk})^2  \Theta(\omega_{k}-\omega_{min})\Theta(\omega_{max}-\omega_k)\Big\rangle,
\end{equation}
can also be used to understand the delocalization.
These density matrices are stored in the files named \ilc{LocalDensityMatrix.dat}, \ilc{SpectralDensityMatrix.dat}, \ilc{AbsoluteDensityMatrix.dat}, and \ilc{ParticipationRatioMatrix.dat}.
The former should be identical to the unit matrix if the full region of all eigenstates is included.
Segments will be defined based on the values of the absoulte value density matrix.
When the value between two sites divided by the diagonal value is less than the value set by \ilo{Threshold} the two sites will be in the same segment. 

\section{Correlation/Autocorralation}
This technique calculates the frequency correlation functions including cross correlations between all sites.
If Autocorralation is specified instead cross-correlations are skipped, which will speed up the calculations significantly for large systems.
The correlation functions are stored up to \ilo{t1max}$\times$\ilo{Timestep} femto seconds.
The first colum in the file \ilc{CorrelationMatrix.dat} contains the time in femtoseconds,
while the following colums contain the correlations functions ordered as in a tridiagonal matrix unless the technique \ilo{Autocorrelation} is specified.
In that case only the autocorrelation functions are calculated and stored with one column per site.
The time-dependent skewness $\langle \omega(t)\omega^2(0) \rangle$, and the time-dependent kutosis $\langle \omega^2(t)\omega^2(0) \rangle$ are stored in the files \ilc{Skewness.dat} and \ilc{Kutosis.dat}, respectively.
These files only contain these higher-order correlation functions for individual sites.
The spectral density (here defined as the real part of the Fourier transform of the time-correlation function) is calculated for each site and stored in the file \ilc{SpectralDensity.dat}.
The first column gives the wavenumber and each remaining column gives the spectral density for a site.
The lineshape function for each site is calculated and stored in the file \ilc{LineshapeMatrix.dat}.
The first column gives the time and the following columns pairwiese give the real and imaginary part of the lineshape function for each site using the temperature given in the input.
The file \ilc{Av\_ExpLineshape.dat} contain the the function $\exp(-g(t))$, where $g(t)$ is the average of the site lineshape functions.
For a collection of uncoupled chrmomphores with identical average energy, this will correspond to the response function including thermal correction.
The first colum is the time in femto seconds, while the two other columns are the real and imaginary parts.

\section{Pop (population transfer)$^{+}$}
The population transfer is calculated between sites.
In general, this is governed by the equation:
\begin{equation}
P_{fi}(t)=\langle |U_{fi}(t,0)|^2 \rangle
\end{equation}
Here $f$ and $i$ are the final and initial sites.
Generally two files are generated.
In \ilc{Pop.dat} average of the population remaining on the initial site ($P_{ii}$) is calculated resulting in a value starting at 1 (all popiulation is on the initial state) and decaying to the equilibrium value $1/N$ (equal population on all states).
In the \ilc{PopF.dat} file a full matrix op all combinations of initial and final states is given.
The columns start start with the first initial state and all possible final states continuing to the second possible initial state and all possible final states.
This file may be very large for large system sizes!
The calculation as default is calculated in the site basis.
A different basis can be defined using the keyword \ilk{Basis}.
The choices \ilo{Local}, \ilo{Average}, and \ilo{Adiabatic} specifies the Local/Site, average exciton, and instantaneous adiabatic basis, respectively.

\section{MCFRET$^{+}$}
The MCFRET part calculates the Excitation Energy Transfer (EET) rate with Multichromophoric Förster Resonance Energy Transfer (MC-FRET) method.\cite{Jang.2004.Phys.Rev.Lett..92.218301,Jang.2007.J.Phys.Chem.B..111.6807,Zhong.2023.J.Chem.Phys..158.064103} The rate is calculated based on the equation:
\begin{equation}\label{eq:17} 
    k=\frac{\rm 2Re}{\hbar^2}\int_{0}^{\infty}dt\ \textrm{Tr}[J^{\textrm {AD}}E^{\textrm D}(t)J^{\textrm {DA}}I^{\textrm A}(t)],
\end{equation}
with $I^{\textrm A}(t)$ and $E^{\textrm D}(t)$ are the absorption time domain matrix and emission time domain matrix. 

The techniques MCFRET-Absorption and MCFRET-Emission can provide the time domain absorption matrix, and time domain emission matrix separately.
For MCFRET-Emission the expectation value of the density matrix is required, which can be obtained with MCFRET-Density.
The technique of MCFRET-Coupling determines the average coupling between segments over a trajectory. The intermediate results are stored in the files:
 \ilc{TD\_absorption\_matrix.dat} for the time domain absorption matrix, \ilc{TD\_emission\_matrix.dat} for the time domain emission matrix, and  \ilc{CouplingMCFRET.dat} for the average couplings. 
The technique of MCFRET-Rate can be used to calculate the EET rate with the input of the time domain absorption matrix, time domain emission matrix, and the coupling to generate the results in steps and introduce possible changes in only some of these steps.  The calculated rate matrix is written in the file {\tt RateMatrix.dat} (in units of ps${-1}$). The MCFRET-Analyse technique calculates the expectation value of the segment energies and the quantum corrected rate matrix, which is stored in 'QC\_RateMatrix.dat' (this includes the Oxtoby correction to give a thermal distribution). The intermediate  rate response is stored in {\tt RateFile.dat}. It is recommended to validate that the response do decay to zero within the chosen simulation time. The coherence decay matrix is stored in {\tt CoherenceMatrix.dat}. The coherence decay should be faster than the transfer rate for MCFRET to be a suitable approximation for the energy transfer.
The MCFRET technique will perform all steps sequentially to calculate the EET rate matrix, but they can be performed sequentially, which may be useful if one for example only change the couplings between segments. 
The MCFRET keyword, thus, corresponds to running the following techniques sequentially: MCFRET-Density, MCFRET-Absorption, MCFRET-Emission, MCFRET-Coupling, MCFRET-Rate, and MCFRET-Analyse.

\section{Redfield}
This technique claculates the Redfield rate matrix (see Eq. 33 of Ref. \citenum{Yang.2000.JCP.113.2823}).
It requires the spectral density as calculated with the Autocorrelation technique and the average Hamiltonian, which can be obtained using the Analyse technique.
The rate equation is then:
\begin{equation}
	k_{ij}=\sum_{n=1} |c_{in}|^2|c_{jn}|^2 C_n(\omega_j-\omega_i),
\end{equation}
For negative frequency differences (upward energy transfer) the spectral sensity is suppressed using the Boltzmann factor: $C_n(-omega)=\exp(-\omega/k_B T)C_n(\omega)$.
Here, $\omega_j$ and $\omega_i$ are the eigen energies of the exciton states $i$ and $j$.
This approximation is applicable in the limit of weak phonon coupling.
That is when the values of the spectral density are small compared to the resonance couplings between sites.
If segments are specified in the input the excitons will be determined only within these segments and only intra-segment rates are determined,
as transfer between weakly coupled segments are likely better described using the MCFRET theory. 
The calculated rate matrix is stored in the file \ilc{RedfieldRates.dat} and the exciton eigenstate information is
stored in the \ilc{RedfieldRates.dat} file.

\section{Dif (diffusion)$^{+}$}
The mean square displacement of excitons are calculated using the positions provided in the Positions file. Periodic boundary conditions are applied assuming a cube box. The box size must be provided in the first column of the Positions file.
Two diffusion properties are calculated. The diffusion of the exciton center is calculated by calculating the center of the excitation given it started on a specific site $j$: $x_{\textrm{ex},i}(t)=\langle \sum_j (x_j(t)-x_i(0)) |U_{ji}(t,0)|^2$ and then determining the mean square displacement as $MSD_{\textrm{ex}}(t)=\langle \sum_i  x_{\textrm{ex},i}(t)^2\rangle$. The other measure is based on the probability of starting on one site and being detected on another site $MSD_{\textrm{site}}(t)=\langle \sum_{ji} (x_j(t)-x_i(0))|U_{ji}(t,0)|^2\rangle$. $MSD_{\textrm{site}}(t)$ is stored in the second column of the Dif.dat output file and $MSD_{\textrm{ex}}(t)$ is stored in the third column. The first column contain the time in femtoseconds. The unit for the mean square displacements is the distance unit provided on the position file (\AA ngstr\"{o}m recommended) squared per femtosecond. 
\section{Ani (anisotropy)}
Not implemented yet (check NISE\_2015)
\section{DOS$^{+}$}
The density of states is calculated using the response function:
\begin{equation}
	I(t)=\langle\textrm{Tr}U(t,0)\rangle\exp(-t/2T_1).
\end{equation}
Both the real and imaginary parts are stored. The Fourier transform is the frequency domain density of states, which is stored in the file DOS.dat. $T_1$ is the lifetime, which is often simply used as an appodization function to smoothen the spectrum.
\section{Absorption}
The linear absorption is calculated using the first-order response function \cite{Duan_2015}
\begin{equation}
	I(t)=\sum_{\alpha}^{x,y,z}\langle\mu_{\alpha}(t)U(t,0)\mu_{\alpha}(0)\rangle\exp(-t/2T_1).
\end{equation}
Both the real and imaginary parts are stored. The Fourier transform is the frequency domain absorption, which is stored in the file Absorption.dat. $T_1$ is the lifetime, which is often simply used as an appodization function to smoothen the spectrum. 
\section{Luminescence}
The luminescence is calculated using the first-order response function
\begin{equation}
	I(t)=\sum_{\alpha}^{x,y,z}\langle\frac{1}{Z}\mu_{\alpha}(t)U(t,0)\exp(-H(0)/k_BT)\mu_{\alpha}(0)\rangle\exp(-t/2T_1).
\end{equation}
Both the real and imaginary parts are stored.
The Fourier transform is the frequency domain luminescence, which is stored in the file \ilc{Luminescence.dat}.
$T_1$ is the lifetime, which is often simply used as an appodization function to smoothen the spectrum.
The Boltzmann term containg the Hamiltonian at time zero ($H$) and the temperature (to be specified in the input) ensure the emission from a termalized population of the excited state ignoring a potential Stoke's shift and effects of vibronic states.
The spectrum is normalized with the partition function ($Z$). 
\section{LD (linear dichroism)}
The linear dichroism is calculated identically to the linear absorption except the average of the absorption in the x and y directions is subtracted from the absorption in the z direction.
This corresponds to a perfect linear dichroism setup, where the molecules are aligned along the z-axis.

\section{CD (circular dichroism)$^{+}$}
The circular dichroism is calculated using the first-order response function
\begin{equation}
	I(t)=\sum_{\alpha}^{x,y,z}\sum_{nm}\langle r_{nm}\mu_{\alpha,n}(t)\times[U(t,0)\mu_{\alpha,m}(0)]\rangle\exp(-t/2T_1).
\end{equation}
Both the real and imaginary parts are stored.
The Fourier transform is the frequency domain absorption, which is stored in the file \ilc{Absorption.dat}.
$T_1$ is the lifetime, which is often simply used as an appodization function to smoothen the spectrum.
Note that the positions of the chromophores must be provided in the \ilk{Positionfile}.
The calculation does \textit{not} account for periodic boundary positions and the positions must at all times along the position trajectory be specified accordingly.

\section{Raman}
The Raman response is calculated as \cite{Torii.2002.J.Phys.Chem.A.106.3281,Shi.2012.J.Phys.Chem.B.116.13821}
\begin{equation}
        I^{VV/VH}(t)=\sum_{a,b,c,d}^{x,y,z}A^{VV/VH}_{abcd}\langle\alpha_{ab}(t)U(t,0)\alpha_{cd}(0)\rangle\exp(-t/2T_1),
\end{equation}
where $A^{VV/VH}_{abcd}$ is the orientational weighting factor as defined in Ref. \citenum{Shi.2012.J.Phys.Chem.B.116.13821}. 
Both the all parallel (VV) and perpendicular (VH) components are calculated.
The calculation require that the transition-polarizability (\ilk{Alphafile}) file is provided with all six independent components stored.
The calculated spectra are stored in the files \ilc{Raman\_VV.dat} and \ilc{Raman\_VH.dat}.

\section{SFG (sum-frequency generation)}
The sum-frequency generation signal is calculated using the first-order response function \cite{Roy.2014.J.Chem.Phys..141.18C502}
\begin{equation}
		I(t)=\sum_{a,b,c}^{x,y,z}A^{SSP/PPP}_{abc}\langle\alpha_{cb}(t)U(t,0)\mu_{a}(0)\rangle\exp(-t/2T_1).
\end{equation}
Both the real and imaginary parts are stored.
The Fourier transform is the frequency domain absorption, which is stored in the files \ilc{SFG\_PPP.dat} and \ilc{SFG\_SSP.dat}.
For PPP $A^{PPP}_{abc}$ is zero except for $a=b=c=1$, where it is one.
For SSP $A^{SSP}_{abc}$ is zero for all components except $a=z$ and $b=c=x,y$, where it is one half.
$T_1$ is the lifetime, which is often simply used as an appodization function to smoothen the spectrum. 

\section{2DIR$^{*}$ (two-dimensional infrared)}
This calculates the two-dimensional infrared spectra assuming coupled three level systems.
The techniques \ilo{GBIR} (ground state bleach), \ilo{SEIR} (stimulated emission), and \ilo{EAIR} (excited state absorption) provides these contributions separetely.
Furthermore, the sum of the ground state bleach and the stimulated emission can be calculated with the \ilo{noEA} technique keyword.
The expressions for the response functions are given in ref. \cite{Jansen.2006.JPCB.110.22910}.

\section{2DSFG (two-dimensional sum-frequency\\ generation)}
Not implemented yet (check NISE\_2015)
 
\section{2DUVvis$^{*}$ (two-dimensional electronic\\ spectroscopy)}
This calculates the two-dimensional infrared spectra assuming coupled two level systems.
The techniques \ilo{GBUVvis} (ground state bleach), \ilo{SEUVvis} (stimulated emission), and \ilo{EAUVvis} (excited state absorption) provides these contributions separately.
Furthermore, the sum of the ground state bleach and the stimulated emission can be calculated with the \ilo{noEAUVvis} technique keyword. 

\section{CG$\_$2DES (coarse-grained two-dimensional electronic spectroscopy)$^{+}$}
This calculates the two-dimensional electronic spectra with the coarse-grained approach.\cite{Zhong.2024.J.Chem.TheoryComput..20.6111}
This is especially efficient for large molecular systems that separate the system into different segments and calculate the 2DES based
on the doorway-window function.
The technique \ilo{CG$\_$2DES$\_$doorway} provides the doorway part of the CG-2DES results.
The techniques \ilo{CG$\_$2DES$\_$window$\_$SE}, \ilo{CG$\_$2DES$\_$window$\_$GB}, and \ilo{CG$\_$2DES$\_$window$\_$EA} provide these
window functions separately. The calculation time is independent of the delay time $t_2$.
\ilo{CG$\_$2DES$\_$combine} can be use to calculate the final spectra using precalculated doorway and window functions.

The response functions governing this technique are provided in ref. \cite{Zhong.2024.J.Chem.TheoryComput..20.6111}.

If a file called \ilc{Waitingtimes.dat} is present all the waiting times given in the file (in fs) will be calculated simultaneously and the response function files will be named \ilc{RparI\_0fs.dat} etc.

\section{2DIRRaman$^{*}$}
This calculates the two-dimensional infrared raman spectra assuming coupled three level systems.
The techniques \ilo{2DIRraman1} (rephasing), \ilo{2DIRraman2} (non-rephasing 1), \ilo{2DIRraman3} (non-repasing 2) provides these contributions separetely.
The rephasing diagram and the non-rephasing diagrams are stored separately as they cannot be directly added.
The expressions for the response functions are given in ref. \cite{vanHengel.2023.J.Chem.Phys..158.064106}.

\section{2DFD (fluorescence detected two-dimensional spectroscopy)}
 The 2DFD spectrum can be calculated in the approximation that all exciton pairs annihilate to produce a single exciton long before fluorescence occur with the \ilk{noEAUVvis} technique.
 The response functions governing this technique are provided in ref. \cite{Kunsel_2018}.
